Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-04-25 ms/spaceness.tex
peter mostly edited until Results

\documentclass[11pt,twoside,lineno]{preprint}
\usepackage{amsmath}
\usepackage[nodisplayskipstretch]{setspace}
\usepackage{makecell} 
\usepackage{changepage}
\setstretch{1.5}
\NeedsTeXFormat{LaTeX2e}
\newcommand*{\articletypename}{PREPRINT}
% Use the documentclass option 'lineno' to view line numbers

@@

\begin{abstract}
Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations. As a result many methods of analyzing genetic data assume that samples are a random draw from a randomly mating population, but are applied to real populations that are structured over space by limited dispersal. Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and genome-wide association studies. We find that continuous space skews the distribution of several common summary statistics, with the most bias occurring when Wright's neighborhood size is less than 100 and sampling is spatially clustered. In turn estimates of population size history based on the site frequency spectrum become biased in the face of limited dispersal. We show that in the context of genome-wide association studies the combination of spatial variation in the environment and limited dispersal creates spurious signals of genetic association with purely environmental phenotypes across much of the genome and that this bias is only partially corrected by regressing out principal components of ancestry. Last we discuss the relevance of our simulation results for inference from genetic variation in real organisms. 
\end{abstract}

@@

\section{Introduction}
The inescapable reality that biological organisms live, move, and reproduce in continuous spatial landscapes has been all but ignored in population genetic models. Indeed, a near universal rule of reproduction is that individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances. This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but classical mathematical models describing it are flawed approximations of the underlying process \citep{Felsenstein1975}.
As a result many studies focus on describing geographic structure as a set of populations connected by migration \citep{Malecot1948,Wright1931}, and analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} assuming these are randomly mating units.

Though discrete models of between-population differentiation account for much of the genetic variation in species, the assumption that populations are randomly mating at some level has important implications for downstream inference of selection and demography. Methods based on the coalescent \citep{Kingman1982} assume that the sampled individuals are a random draw from a randomly mating population. If dispersal or mate selection is limited by space these assumptions are violated -- nearby individuals will be more closely related than an average random pair and drawing multiple samples from the same point on the landscape will represent a biased sample of the genetic variation present in the whole population. All downstream inferences drawn from the patterns of relationship inferred among sampled individuals will be subject to bias, but the extent depends on the degree of variation in ancestry across the landscape.

Our understanding of how many common summary statistics respond to variation in dispersal distance is limited. Tajima's $D$, for example, is often interpreted as reflecting evidence of selection or past population size changes (CITE \cjb{CJ: what's a good citation here?}). However this statistic is essentially a summary of the site frequency spectrum, which itself reflects variation in branch lengths of the underlying genealogy of sampled individuals. Because the spatial extent of dispersal and mate selection determines the rate of drift in a continuously distributed population \citep{Maruyama1972}, these population genetic summaries should also reflect variation in the scale of spatial interactions both over time and across populations. Similarly, the distribution of tract lengths of identity by state among individuals is often interpreted either through the lens of historic demography \citep{Harris2013} or selection \citep{Garud2015}. While that is so this distribution also contains abundant information about dispersal and mate selection \citep{Ringbauer2017,Baharian2016}. We are particularly keen on examining how such summaries will be affected by models that incorporate continuous space.

A related issue is the spatial distribution of sampling effort. Though individuals in a random-mating population are exchangeable, those in a continuously structured population are not. If sampling is clustered at a series of points on the landscape, as is common in empirical studies because of the practical difficulties of sample collection, samples drawn from a spatially structured population may partially reflect spatial biases in the sampling process rather than in the underlying population. In addition, range edges have been observed to create complex patterns of heterozygosity in stepping-stone simulations \citep{Neel2013,Shirk2014}, but the effects of this process on many downstream inference procedures is unknown. \cjb{must be more out there on sampling -- any recs for papers to read+cite?}

The issue of fine-scale structure for putatively "within-population" analyses may have important implications for genome-wide association studies as well. This is because the genome-wide allele frequency variation caused by isolation by distance is similar to that expected when many sites in the genome are associated with a given phenotype \citep{Bulik-Sullivan2015}. When population structure and the environment covary across space genome-wide association studies (GWAS) may incorrectly infer phenotypic association at many SNPs that have differentiated by drift \citep{Mathieson2012}. This may be a particularly important effect when using GWAS to study selection on polygenic traits. Indeed two recent studies found that previous evidence of polygenic selection on human height in Europe was confounded by subtle population structure \citep{Sohail2018,Berg2018}, suggesting that existing methods to correct for population structure in GWAS are insufficient. However we have little quantitative idea of the population and environmental parameters that can be expected to lead to biases in GWAS. As the scale of sequence data now available for many species allows inference of increasingly fine-scale patterns of selection and demography, understanding how and when subtle spatial structure is likely to bias results is an important task for population genetics.

Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness. The model scales to chromosome-scale alignments across tens of thousands of individuals and outputs the full genealogy and recombination history of all final-generation individuals. We use this simulation to test how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed assuming discrete population models. We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum. Lastly we examine the impacts of evolution in continuous space on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.

@@

\section{Materials and Methods}
\label{sec:materials:methods}

\subsection{Modeling Evolution in Continuous Space}
The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1942,Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. A key finding of Wright's model is that many important aspects of continuous populations can be described in terms of "neighborhood size" -- the number of potential mates for an individual in a given generation, defined as $4\pi\sigma^2d$, where $\sigma$ is the average dispersal distance and $d$ is population density. \cite{Maruyama1972} found that the rate of decline in genetic diversity in a 2-dimensional continuous population approaches the random mating expectation when $d\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

@@

One method for modeling continuous populations is then to assume the existence of a grid of discrete randomly-mating populations connected by migration, which prevents clustering by forcing all regions to be occupied in every generation. Among many other important results drawn from this class of "lattice" or "stepping stone" models, \cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{st}$) against the logarithm of spatial distance is an estimate of neighborhood size. Though good approximations of continuous structure given high dispersal, these models are not truly continuous, force a uniform realized population density across landscapes, and limit investigation of spatial structure below the level of the deme. An alternative method is to model the geographic spread of ancestry backwards in time through a diffusion approximation -- an approach that has recently made significant progress in modeling both dispersal and demographic parameters \citep{Barton2010,Kelleher2014,Ringbauer2017,Ringbauer2018}.

We took a direct approach to the clustering problem of classical forward-time models by incorporating density dependence into an individual-based model similar to the analytic models developed by Wright and Malécot. By scaling the probability of survival in each timestep to local population density we shift reproductive output towards regions of low-density and prevent populations from clustering. A similar approach was taken previously by \citep{Doebeli2003} who used an individual based model with continuous space and density dependent fitness to study the probability of speciation along continuous environmental gradients. However to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci as the unit of analysis, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.
\subsection{A Forward-Time Model of Evolution in Continuous Space}

We implemented our model using the non-Wright-Fisher module in the program SLiM v3.0 \citep{Haller2019}. Each time step consists of three stages: reproduction, dispersal, and competition. To reduce the parameter space we use the same parameter, denoted $\sigma$, to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ as applied in our dispersal step is approximately equal to the mean parent-offspring distance.

At the beginning of the simulation individuals are distributed randomly on a continuous landscape. Mates are selected proportional to distances among individuals weighted by a Gaussian function with mean $1/(2\pi\sigma^2)$ and standard deviation $\sigma$. Wright's \cite{Wright1943} "Neighborhood Size", defined as $4\pi\sigma^2 K$ where $K$ is the population density, is then the approximate number of individuals available for mating in our simulation. The number of offspring is drawn from a Poisson distribution with mean $1/L$, where $L$ is the average lifespan. If new offspring are produced, we simulate dispersal by taking two draws from a normal distribution with mean 0 and standard deviation $\sigma$ and adding these to the x and y coordinates of the first parent.

Density-dependent competition is simulated by adjusting the probability of survival of each individual proportional to the local density of individuals up to a distance $3\sigma$ away, scaled according to the same Gaussian function used for mate selection. Given a per-unit carrying capacity $K$ and average fecundity $F$, the probability of survival $d$ for individual $i$ after adjustment for density is:

\begin{equation}
    d_{i}=min(0.95,\frac{1}{1+\rho*\sum_{i}{G(x)}})
\end{equation}
where 
\begin{equation}
    \rho = F/((1+F)*K)
\end{equation} 
and $G(x)$ is the Gaussian function used to weight distances to other individuals.

A major challenge in all spatial models is dealing with range edges. When local population density is used to model competition, edge or corner populations can be assigned artificially high fitness values because they lack neighbors within their interaction radius but outside the bounds of the simulation. We approximate a decline in habitat suitability near edges by decreasing the probability of survival proportional to the square root of distance to edges in units of $\sigma$. The final probability of survival for individual \textit{i} is then

\begin{equation}
    s_{i}=d_{i} min(1,\sqrt{x_{i}/\sigma})
    min(1,\sqrt{y_{i}/\sigma})\\
    min(1,\sqrt{(x_{max}-x_{i})/\sigma})
    min(1,\sqrt{(y_{max}-y_{i})/\sigma)}
\end{equation}

where $x$ and $y$ are spatial coordinates. In practice this creates a one $\sigma$ buffer around the edge of the map in which fitness is depressed.

To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness, we  also implemented a second version of the SLiM simulation in which mates are selected at random and offspring disperse to a random location on the landscape. We refer to these two models as the "spatial" and "random mating" models. For all simulations the full genealogy and recombination history of final-generation individuals were stored as tree sequences \citep{Kelleher2018}. Scripts for figures and analyses are available at https://github.com/petrelharp/spaceness.

We ran 400 simulations for the spatial and random-mating SLiM models on a square landscape of size 50 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), average lifetime $L=4$, genome size $1\times10^{8}$, recombination rate $1\times10^{-9}$, and drawing $/sigma$ values from a uniform distribution bounded by 0.2 and 4. To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM and later applied mutations to the coalesced tree sequence with msprime's 'mutate( )' function. Because msprime assumes that timesteps are in units of generations, we divided the per-generation rate of $1\times10^{-8} mut/site/gen$ by the generation time estimated for each value of $\sigma$ (see 'Genealogical Parameters' below) to convert the rate to units of mutations/site/timestep. We then verified that this procedure produced the correct number of mutations by comparing a subset of simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime. Simulations were run for 1.6 million timesteps (approximately 30N generations), or until all extant individuals shared a common ancestor within the simulation (i.e. the tree sequence had coalesced). \ak{maybe worth including a table with some basic runtime results in the supplement?}

\subsection{Genealogical Parameters}
Because basic elements of our simulations like census population size and generation time are not directly parameterized in our model (instead they emerge from the interaction of parameters including average population density, average lifespan, and the spatial scale of interaction), we first monitored a range of genealogical parameters in all simulations. We recorded the census population size and estimated the variance in number of offspring and generation time for all simulations. To estimate generation times we stored the age of all parents for 200 timesteps after a 100 generation burn-in and took the mean. To estimate variance in offspring we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.

\subsection{Sampling}
Our model records the genealogy and sequence variation of the complete population, but in real data genotypes are only observed from a relatively small number of sampled individuals. We modeled three sampling strategies similar to common data collection methods in empirical genetic studies (Figure \ref{fig:samplemap}). "Random" sampling selects individuals at random from across the full landscape, "point" sampling selects individuals proportional to their distance from four equally spaced points on the landscape, and "midpoint" sampling selects individuals in proportion to their distance from the middle of the landscape. Downstream analyses were repeated across all sampling strategies.

\begin{figure}[htbp]
\centering
\includegraphics{figures/sampling_maps.pdf}
\caption{Example sampling maps for 60 individuals on a 50x50 landscape under midpoint, point, and random sampling strategies.}
\label{fig:samplemap}
\end{figure}

\subsection{Summary Statistics}
We calculated the site frequency spectrum and a set of 18 summary statistics (Table \ref{table:sumstats}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{is}$), and Tajima's D, as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, which we summarized as the correlation coefficient between $log10(spatial distance)$ and the proportion of identical base pairs among individuals.

Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among samples. The full distribution of lengths of IBS tracts for each pair of individuals was first calculated with a custom python function (https://github.com/petrelharp/spaceness/blob/master/scripts/slimtools.py). We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $1e6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.

We then estimated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution. Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e. most negative) when dispersal is low. The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks so should reflect recent demographic events \citep{Ringbauer2017}.

@@

\subsection{Demographic Modeling}
We fit single-population demographic models to the site frequency spectra of 20 individuals from each spatial SLiM simulation with the program Stairwayplot \citep{Liu2015}. This analysis was replicated across random, point, and midpoint sampling strategies. Site frequency spectra used for input data were calculated in scikit-allel \citep{Miles2017}, and 100 bootstrap replicates were generated for each simulation by resampling over sites. Models were fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $Ne$ per generation was used to represent the output of each simulation.

In preliminary runs we found that inferred population histories were relatively variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal. To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $\Theta_{\pi}$) and fit stairwayplot models using the same script as for our spatial models. All coalescent simulations were performed using msprime \citep{Kelleher2016}. We then calculated the standard deviation of inferred $N_{e}$ in each stairwayplot model to summarize the degree of fluctuation around the simulated population size, and asked if standard deviations were higher in spatial relative to coalescent models with a one-tailed t-test.

>>>>

\documentclass[11pt,twoside,lineno]{preprint}
\usepackage{amsmath}
\usepackage[nodisplayskipstretch]{setspace}
\usepackage{makecell} 
\usepackage{changepage}
\usepackage{afterpage}
\setstretch{1.5}
\NeedsTeXFormat{LaTeX2e}
\newcommand*{\articletypename}{PREPRINT}
% Use the documentclass option 'lineno' to view line numbers

@@

\begin{abstract}
    Real geography is continuous,
    but standard models in population genetics are based on discrete, well-mixed populations. 
    As a result many methods of analyzing genetic data assume that samples are a random draw from a well-mixed population, 
    but are applied to real populations that have significant geographic structure.
    Here we use simulations of populations living in continuous geography 
    to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and genome-wide association studies. 
    We find that several common summary statistics have distributions that differ substantially from that seen in well-mixed populations, 
    especially when Wright's neighborhood size is less than 100 and sampling is spatially clustered. 
    Standard estimates of population size history based on the site frequency spectrum become biased with lower dispersal. 
    We also show that the combination of spatially autocorrelated environments and limited dispersal
    can cause genome-wide association studies to identify numerous spurious signals of genetic association 
    with purely environmentally determined phenotypes,
    and that this bias is only partially corrected by regressing out principal components of ancestry. 
    Last, we discuss the relevance of our simulation results for inference from genetic variation in real organisms. 
\end{abstract}

@@

\plr{note: address confounding of $\sigma$ and neighborhood size in disco}

\plr{note: conclusions are that Ne(t) is not very biased}

%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

The inescapable reality that biological organisms live, move, and reproduce in continuous geography is usually omitted from population genetic models. 
Indeed, a near universal rule of reproduction is that individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances. 
This pattern of ``isolation by distance'' \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, 
but classical mathematical models describing it \citep{Malecot1948} are flawed approximations of the underlying process \citep{Felsenstein1975,barton2002neutral}.
Mathematical difficulties have made further development of population models in continuous space difficult,
so most models describe geographic structure as a set of populations connected by migration \citep[e.g.,][]{Wright1931,epperson2003geographical}, 
and most empirical papers analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000},
effectively assuming these are randomly mating units.

The assumption that populations are ``well-mixed'' has important implications for downstream inference of selection and demography. 
Methods based on the coalescent \citep{Kingman1982,wakeley2005coalescent} assume that the sampled individuals are a random draw from a well-mixed population
that is much larger than the sample \citep{wakeley2003gene}.
The key assumption is actually that the individuals of each generation are \emph{exchangeable},
so that there is no correlation between the fate or fecundity of a parent and that of their offspring \citep{Huillet2011}.
If dispersal or mate selection is limited by geographic proximity, this assumption can be violated in many ways.
For instance, if mean viability or fecundity is spatially autocorrelated,
then limited geographic dispersal will lead to parent--offspring correlations.
Futhermore, nearby individuals will be more closely related than an average random pair,
so drawing multiple samples from the same area of on the landscape will represent a biased sample of the genetic variation present in the whole population. 
It has long been appreciated that this model misspecification subjects downstream inferences to bias, 
but the extent and nature of these effects remain largely uninvestigated.

For instance, nonzero values of Tajima's $D$ are often interpreted as reflecting evidence of selection or past population size changes \citep{tajima1989statistical}.
This statistic is a summary of the site frequency spectrum, which itself reflects variation in branch lengths and tree structure of the underlying genealogies of sampled individuals.
Geographically limited mate choice distorts the distribution of these genealogies \citep{Maruyama1972},
which can affect the value of Tajima's $D$.
% Because the spatial extent of dispersal and mate selection determines the rate of drift in a continuously distributed population 
% these population genetic summaries should also reflect variation in the scale of spatial interactions both over time and across populations. 
Similarly, the distribution of tract lengths of identity by state among individuals contains information about not only
historical demography \citep{Harris2013,ralph2013geography} and selection \citep{Garud2015}, but also dispersal and mate choice \citep{Ringbauer2017,Baharian2016}. 
We are particularly keen to examine how such summaries will be affected by models that incorporate continuous space,
both to evaluate the assumptions underlying existing methods, but also to identify where the most promising signals of geography lie.

A related issue is the spatial distribution of sampling effort --
since nearby individuals are more likely to be closely related than distant ones,
observed patterns of relatedness are expected to depend on the geographic sampling scheme,
perhaps strongly.
In addition, range edges have been observed to create complex patterns of heterozygosity in stepping-stone simulations \citep{Neel2013,Shirk2014}, 
but the effects of this process on many downstream inference procedures is unknown.
\cjb{must be more out there on sampling -- any recs for papers to read+cite?}

The issue of fine-scale geographic structure may have particularly important implications for genome-wide association studies (GWAS). 
This is because many phenotypes of interest have strong geographic differences due to the (nongenetic) influence of environmental or socioeconomic factors,
which can therefore show spurious correlations with spatially patterned allele frequencies \citep{Bulik-Sullivan2015,Mathieson2012}. 
This may be particularly important for the study of selection on polygenic traits,
whose heritable genetic components are determined by many loci of weak effect,
those hardest to disentangle from spurious correlations. 
Indeed, two recent studies found that previous evidence of polygenic selection on human height in Europe was confounded by subtle population structure \citep{Sohail2018,Berg2018}, suggesting that existing methods to correct for population structure in GWAS are insufficient. However we have little quantitative idea of the population and environmental parameters that can be expected to lead to biases in GWAS. As the scale of sequence data now available for many species allows inference of increasingly fine-scale patterns of selection and demography, understanding how and when subtle spatial structure is likely to bias results is an important task for population genetics.

To study this, we have implemented an individual-based model in continuous geography 
that incorporates overlapping generations, local dispersal of offspring, and density-dependent survival. 
We simluate chromosome-scale genomic data in tens of thousands of individuals, and output the full genealogy and recombination history of all final-generation individuals. 
We use these simulations to test how sampling strategy interacts with geographic population structure to cause systematic variation in population genetic summary statistics 
typically analyzed assuming discrete population models. 
We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum. 
Lastly, we examine the impacts of evolution in continuous space on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}
\label{sec:materials:methods}

%%%%%%%%%%%%%
\subsection{Modeling Evolution in Continuous Space}

\plr{PETER TO REVISIT THIS SECTION}

The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1942,Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. A key finding of Wright's model is that many important aspects of continuous populations can be described in terms of ``neighborhood size'' -- the number of potential mates for an individual in a given generation, defined as $4\pi\sigma^2d$, where $\sigma$ is the average dispersal distance and $d$ is population density. \cite{Maruyama1972} found that the rate of decline in genetic diversity in a 2-dimensional continuous population approaches the random mating expectation when $d\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

@@

One method for modeling continuous populations is then to assume the existence of a grid of discrete randomly-mating populations connected by migration, which prevents clustering by forcing all regions to be occupied in every generation. Among many other important results drawn from this class of ``lattice'' or ``stepping stone'' models, \cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{ST}$) against the logarithm of spatial distance is an estimate of neighborhood size. Though good approximations of continuous structure given high dispersal, these models are not truly continuous, force a uniform realized population density across landscapes, and limit investigation of spatial structure below the level of the deme. An alternative method is to model the geographic spread of ancestry backwards in time through a diffusion approximation -- an approach that has recently made significant progress in modeling both dispersal and demographic parameters \citep{Barton2010,Kelleher2014,Ringbauer2017,Ringbauer2018}.

We took a direct approach to the clustering problem of classical forward-time models by incorporating density dependence into an individual-based model similar to the analytic models developed by Wright and Malécot. By scaling the probability of survival in each timestep to local population density we shift reproductive output towards regions of low-density and prevent populations from clustering. A similar approach was taken previously by \citep{Doebeli2003} who used an individual based model with continuous space and density dependent fitness to study the probability of speciation along continuous environmental gradients. However to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci as the unit of analysis, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

%%%%%%%%%%%%
\subsection{A Forward-Time Model of Evolution in Continuous Space}

We simulated populations using the non-Wright-Fisher module in the program SLiM v3.0 \citep{Haller2019}. 
Each time step consists of three stages: reproduction, dispersal, and mortality. 
To reduce the parameter space we use the same parameter, denoted $\sigma$, 
to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
As in previous work \citep{Wright1943,Ringbauer2017}, $\sigma$ as applied in our dispersal step is equal to the mean parent-offspring distance.  
A key parameter we report below is
\citeauthor{Wright1943} citeyearpar{Wright1943} ``neighborhood size'', 
defined to be $N_W = 4\pi\sigma^2 \rho$ where $\rho$ is the population density.
This the approximate number of individuals available for mating in our simulation.

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ 
who disperse a random, Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates, reflected to stay within the species range.
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2 / 2 \sigma^2)$.

To maintain a stable population, mortality increases with local population density.
To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$,
where $g$ is the Gaussian density with mean zero and standard deviation $\sigma$.
Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
where $d_{ij}$ is the distance between individuals $i$ and $j$ and the sum is over all neighbors within distance $3 \sigma$.
Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
Then, given a per-unit carrying capacity $K$, the probability of survival until the next time step for individual $i$ is
\begin{equation}
    p_i \min(0.95, \frac{1}{1 + n_i / (K (1+L)} .
\end{equation}
We chose this functional form so that the equilibrium population density per unit area is around $K$,
and the mean lifetime is around $L$.

An important step in creating any spatial model is dealing with range edges. 
Because local population density is used to model competition, edge or corner populations can be assigned artificially high fitness values 
because they lack neighbors within their interaction radius but outside the bounds of the simulation.
We approximate a decline in habitat suitability near edges by decreasing the probability of survival proportional to the square root of distance to edges in units of $\sigma$. 
The final probability of survival for individual $i$ is then
\begin{equation}
    s_{i} = p_{i} 
        \min(1,\sqrt{x_{i}/\sigma})
        \min(1,\sqrt{y_{i}/\sigma})\\
        \min(1,\sqrt{(W - x_{i})/\sigma})
        \min(1,\sqrt{(W - y_{i})/\sigma)}
\end{equation}
where $x_i$ and $y_i$ are the spatial coordinates of individual $i$, and $W$ is the width (and height) of the square habitat.
This buffer roughly counteracts the increase in fitness individuals close to the edge would otherwise have.

To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness, 
we also implemented simulations identical to those above except that mates are selected uniformly random from the population,
and offspring disperse to a uniform random location on the landscape. 
We refer to this model as the ``random mating'' model, in contrast to the first, ``spatial'' model.

We stored the full genealogy and recombination history of final-generation individuals as tree sequences \citep{Kelleher2018},
as implemented in SLiM \citep{haller2019treesequence}. 
Scripts for figures and analyses are available at \url{https://github.com/petrelharp/spaceness}.

We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$ 
with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
average lifetime $L=4$, 
genome size $10^{8}$, 
recombination rate $10^{-9}$, 
and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM 
and later applied mutations to the tree sequence with msprime's \texttt{mutate} function \citep{Kelleher2016}. 
Because msprime applies mutations proportionally to elapsed time,
we divided the mutation rate of $10^{-8}$ mutations per site per generation
by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time.
(We verified that this procedure produced the correct number of mutations by comparing to a subset of simulations with SLiM-generated mutations, which are applied only at meiosis.)
Simulations were run for 1.6 million timesteps (approximately $30N$ generations), or until all extant individuals shared a common ancestor within the simulation (i.e., the tree sequence had coalesced). 
\ak{maybe worth including a table with some basic runtime results in the supplement?}

%%%%%%%%%%%%%
\subsection{Demographic Parameters}

Our demographic model includes paramters for population density ($K$),
mean life span ($L$), and dispersal distance ($\sigma$).
However, nonlinearity of local demographic stochasticity causes
actual realized averages of these demographic quantitites to deviate from the specified values
in a way that depends on the neighborhood size.
Therefore, to properly compare to theoretical expectations,
we empirically calculated these demographic quantities in simulations. 
We recorded the census population size in all simulations. 
To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
after a 100 generation burn-in, and took the mean. 
To estimate variance in offspring number, 
we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.
All calculations were performed with information recorded in the tree sequence,
using pyslim \citetext{\url{https://github.com/tskit-dev/pyslim}}.

%%%%%%%%%%%%%
\subsection{Sampling}
Our model records the genealogy and sequence variation of the complete population, but in real data, genotypes are only observed from a relatively small number of sampled individuals. We modeled three sampling strategies similar to common data collection methods in empirical genetic studies (Figure \ref{fig:samplemap}). ``Random'' sampling selects individuals at random from across the full landscape, ``point'' sampling selects individuals proportional to their distance from four equally spaced points on the landscape, and ``midpoint'' sampling selects individuals in proportion to their distance from the middle of the landscape. Downstream analyses were repeated across all sampling strategies.

\begin{figure}[htbp]
\centering
\includegraphics{figures/sampling_maps.pdf}
\caption{Example sampling maps for 60 individuals on a 50$\times$50 landscape for midpoint, point, and random sampling strategies, respectively.}
\label{fig:samplemap}
\end{figure}

%%%%%%%%%%%%%
\subsection{Summary Statistics}
We calculated the site frequency spectrum and a set of 18 summary statistics (Table \ref{table:sumstats}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. 
Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{IS}$), and Tajima's $D$, 
as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, 
which we summarized as the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals.

Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among samples. The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
%\url{https://github.com/petrelharp/spaceness/blob/master/scripts/slimtools.py}. 
We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $1e6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.

We then estimated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution. Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e. most negative) when dispersal is low. The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks so should reflect recent demographic events \citep{chapman2002effect}.

@@

%%%%%%%%%%%
\subsection{Demographic Modeling}

We fit single-population demographic models to the site frequency spectra of 20 individuals from each spatial SLiM simulation with the program Stairwayplot \citep{Liu2015}. This analysis was replicated across random, point, and midpoint sampling strategies. Site frequency spectra used for input data were calculated in scikit-allel \citep{Miles2017}, 
and 100 bootstrap replicates were generated for each simulation by resampling over sites. 
\plr{what were bootstrap replicates used for? if anything, need to say more precisely what you mean here}
Models were fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $N_e$ per generation was used to represent the output of each simulation.

In preliminary runs we found that inferred population histories were highly variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal. To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $\Theta_{\pi}$) and fit stairwayplot models using the same script as for our spatial models. All coalescent simulations were performed using msprime \citep{Kelleher2016}. We then calculated the standard deviation of inferred $N_{e}$ in each stairwayplot model to summarize the degree of fluctuation around the simulated population size, and asked if standard deviations were higher in spatial relative to coalescent models with a one-tailed t-test.

%%%%%%%%%%%
\subsection{Association Studies}
